---
title: "Heart Disease ML practice"
author: "Richard Lim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r library, include= FALSE}
library(tidyverse)
library(caret)
library(xgboost)
library(janitor)
library(data.table)
library(DataExplorer)
library(naniar)
library(gridExtra)
library(party)
library(corrplot)
```

```{r load_r, echo = FALSE}
getwd()
setwd("C:/Users/dchan/Downloads")
heart <- fread("heart.csv") %>% clean_names()
```

# Data Variable Dictionary

1. age - age in years
2. sex - (1 = male; 0 = female)
3. cp - chest pain type
  + 0: Typical angina: chest pain related decrease blood supply to the heart
  + 1: Atypical angina: chest pain not related to heart
  + 2: Non-anginal pain: typically esophageal spasms (non heart related)
  + 3: Asymptomatic: chest pain not showing signs of disease
4. trestbps - resting blood pressure (in mm Hg on admission to the hospital) anything above 130-140 is typically cause for concern
5. chol - serum cholestoral in mg/dl
  + serum = LDL + HDL + .2 * triglycerides
  + above 200 is cause for concern
6. fbs - (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false)
  + '>126' mg/dL signals diabetes
7. restecg - resting electrocardiographic results
  + 0: Nothing to note
  + 1: ST-T Wave abnormality
    + can range from mild symptoms to severe problems
    + signals non-normal heart beat
  + 2: Possible or definite left ventricular hypertrophy
    + Enlarged heart's main pumping chamber
8. thalach - maximum heart rate achieved
9. exang - exercise induced angina (1 = yes; 0 = no)
10. oldpeak - ST depression induced by exercise relative to rest looks at stress of heart during excercise unhealthy heart will stress more
11. slope - the slope of the peak exercise ST segment
  + 0: Upsloping: better heart rate with excercise (uncommon)
  + 1: Flatsloping: minimal change (typical healthy heart)
  + 2: Downsloping: signs of unhealthy heart
12. ca - number of major vessels (0-3) colored by flourosopy
  + colored vessel means the doctor can see the blood passing through
  + the more blood movement the better (no clots)
13. thal - thalium stress result
  + 1,3: normal
  + 6: fixed defect: used to be defect but ok now
  + 7: reversable defect: no proper blood movement when excercising
14. target - have disease or not (1=yes, 0=no) (= the predicted attribute)

# EDA (Explanatory Data Analysis)

```{r EDA}
# structure of data
cat("Heart dataset has ", dim(heart)[1], "rows and ", dim(heart)[2], "columns")
str(heart)
# glimpse(heart)
summary(heart)
```

```{r EDA2}
# variables' datatype
lapply(heart, typeof) %>% data.frame() %>% t() %>% as.data.frame()
```

```{r EDA3}
# missing values
p1 <- plot_missing(heart, ggtheme = theme_bw())
p2 <- gg_miss_var(heart)
grid.arrange(p1,p2, ncol = 2)
```

# Data Engineering

```{r engineering}
heart <- 
  heart %>% mutate(
  sex = ifelse(sex == 1, "M", "F"),
  cp = case_when(
    cp == 0 ~ "Typical angina",
    cp == 1 ~ "Atypical angina",
    cp == 2 ~ "Non-anginal pain",
    cp == 3 ~ "Asymptomatic"
  ) %>% as.factor(),
  fbs = ifelse(fbs == 1, "> 120", "< 120"),
  restecg = case_when(
    restecg == 0 ~ "Nothing",
    restecg == 1 ~ "Abnormality",
    restecg == 2 ~ "Hypertrophy"
  ) %>% as.factor(),
  exang = ifelse(exang == 1, "Y","N"),
  slope = case_when(
    slope == 0 ~ "Upsloping",
    slope == 1 ~ "Flatsloping",
    slope == 2 ~ "Downsloping"
  ) %>% as.factor()
    )

heart <- heart %>% mutate_if(is.character, as.factor)

lapply(heart, typeof) %>% data.frame() %>% t() %>% as.data.frame()

colSums(is.na(heart)) %>% data.frame()
plot_intro(heart, ggtheme = theme_bw())
```

```{r model}
set.seed(1000)
model <- createDataPartition(heart$target, p = 0.7, list = F)
train <- heart %>% dplyr::slice(model)
test <- heart %>% dplyr::slice(-model)

rf_model <- train(factor(target) ~., data = heart, method = "rf", importance = T)

ggplot(varImp(rf_model)) + geom_label(aes(label = paste0(round(Importance),"%"))) + theme_bw() + labs(title = "Random Forest") + theme(plot.title = element_text(hjust = 0.5))

test <- test %>% mutate(rf_pred = predict(rf_model, test))

confusionMatrix(factor(test$target), test$rf_pred)

heart <- heart %>% mutate(rf_pred = predict(rf_model, heart))

confusionMatrix(factor(heart$target), heart$rf_pred)
```

```{r model_2}
plot(rf_model$finalModel)
```

