---
title: "At-risk Student Project"
author: "Richard Lim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    keep_md: true
    toc: true
    toc_width: 10
    toc_float: true
    code_folding: hide
---

# 1. Objective of analysis

- Objective: Support academic advising center by building prediction models for who to advise.

- $advsing\_new.csv$ & $student\_new.csv$ are separated but relational datasets which have joining keys.

- Implement Data Mining to figure out which predictor variables to forcast whether student are graduated or not

- Examine if intervention from advising center will help the sucess of students (Graduation)

```{r load_library, include=FALSE}
# Load the Packages
library(data.table)
library(janitor)
library(tidyverse)
library(DataExplorer)
library(naniar)
library(skimr)
library(plotly)
library(corrplot)
library(DT)
library(caret)
library(stringi)
library(lubridate)
library(gridExtra)
library(ggrepel)
library(caret)
library(corrplot)
library(rattle)
```


```{r load_data, include = FALSE}
# Load the data
student_new <- fread('https://raw.githubusercontent.com/BYUIDSS/advising_studentriskW19/master/raw_data/student_new.csv?token=AKKUOQEY5H2TZBA4BKOWGDS6L74DW',stringsAsFactors = F)
advising_new <- fread('https://raw.githubusercontent.com/BYUIDSS/advising_studentriskW19/master/raw_data/advising_new.csv?token=AKKUOQGHOCQVMGHDPIMY5JK6L74CE',stringsAsFactors = F)
```

# 2. EDA (Expolatory Data Analysis)

## a. Scales of data

```{r attr_data}
# ncol() | nrow()
cat("The student_new has", dim(student_new)[1], "rows and", dim(student_new)[2], "columns.\nThe advisng_new has",dim(advising_new)[1], "rows and", dim(advising_new)[2], "columns.")
```

There are 52 variables to inspect if it might be a good source to predict who is going to sucess on school or not.


## b. Type of variables 

```{r summary_data}
lapply(c(student_new, advising_new), typeof) %>% data.frame() %>%  t() %>%  as.data.frame() 
```



## c. How many missing values for each dataset?

There are observations which are represented as $NA$, but it is recorded as empty string instead of it. Therefore, it needs to be corrected and Identified as missing value.

```{r missing_transform}
student_new <- sapply(student_new, function(x) {ifelse(is.character(x) & nchar(x) == 0, NA, x)}) %>% data.frame(stringsAsFactors = F) # Careful to not use as.data.frame()
advising_new <- sapply(advising_new, function(x) {ifelse(is.character(x) & nchar(x) == 0, NA, x)}) %>% data.frame(stringsAsFactors = F)
```

```{r missing_value, fig.show = 'hide'}
missing_student <- plot_missing(student_new, missing_only = T, theme_config = list(legend.position = "right"), ggtheme = theme_light(), title = "student_new missing") %>% ggplotGrob()
missing_advising <- plot_missing(advising_new, theme_config = list(legend.position = "right"), ggtheme = theme_light(), title = "advising_new missing") %>% ggplotGrob()
```

```{r missing_result, fig.height= 6}
grid.arrange(missing_student, missing_advising, ncol = 1)
```

- It appears that $pathway\_id$, $deferred\_term$, $deferred\_term\_session\_order$, $track\_override$ are the variables which is not suitable for prediction modeling.

- The missing plot per variable can be created with $naniar$ Package as well below

- There is $gg\_miss\_var$ plot which indicates how much proportion of missing values in $nanair$ pacakge below. 

```{r nanair_1}
gmv1 <- naniar::gg_miss_var(student_new, show_pct = T)
gmv2 <- naniar::gg_miss_var(advising_new, show_pct = T)
grid.arrange(gmv1, gmv2, nrow = 1)
```

- Also, it can be visualized with other way shown below.

```{r nanair_2, fig.height= 10}
m1 <- naniar::vis_miss(student_new, warn_large_data = F) + labs(title = "Student_new missing")
m2 <- naniar::vis_miss(advising_new, warn_large_data = F) + labs(title = "Advsing_new missing")
grid.arrange(m1, m2, ncol = 1)
```

### - Cleansing the column (Variable) names with Janitor Package

In order to access column names more conviniently, I used $Janitor$ Package to cleanse the name of columns to code in R-friendly way.

```{r clean_names}
# Clean the name of variables (lower.case & '_' connection)
student_new <- student_new %>% clean_names()
advising_new <- advising_new %>% clean_names()
```


### - Switching original continous features to numeric value

In the process of changing non-typed observations into NA, all variables' type has been changed into $character$. GPA, ACT Score, duration and Credits related variables must be treated as Numeric values. 

```{r numeric}
student_new <- student_new %>% mutate_at(colnames(student_new) %>% grep("gpa|act|credits", ., value =T), as.numeric)
advising_new <- advising_new %>% mutate(total_meeting_duration = total_meeting_duration %>% as.numeric())

t1 <- student_new %>% select_if(is.numeric) %>% colnames() %>% data.frame() 
t2 <- advising_new %>% select_if(is.numeric) %>% colnames() %>% data.frame()
rbind(t1, t2) %>% mutate(type = "Numeric")
```

### - DataExplorer Package 

```{r DE,  fig.show = 'hide'}
# DataExplor package - numeric variable & categorical variable & row & column
t1 <- DataExplorer::introduce(student_new) %>% t() %>% as.data.frame() %>% rownames_to_column("attribute")
t2 <- DataExplorer::introduce(advising_new) %>% t() %>% as.data.frame() %>% rownames_to_column("attribute")
p1 <- DataExplorer::plot_intro(student_new, title = "student data", ggtheme = theme_bw())
p2 <- DataExplorer::plot_intro(advising_new, title = "advising data", ggtheme = theme_bw())
```

```{r DE_result,  fig.height= 6, fig.width=  10}
grid.arrange(p1, tableGrob(t1, rows=NULL), tableGrob(t2, rows=NULL), p2,  ncol = 2, top = "Attributes of datasets")
```



# 3. Feature Exploration

## a. The number of student for each dataset

```{r num_student}
# The number of student
cat("1. There are ",student_new %>% distinct(identifier) %>% nrow(), "students for student_new dataset.\n2. There are ",advising_new %>% distinct(identifier) %>% nrow(), "students for advising_new dataset.\n3. There are ", student_new %>% distinct(identifier) %>% nrow() - advising_new %>% distinct(identifier) %>% nrow(), "students who has not visited advising center for the periods")

# unique(student_new$identifier) %>% length()

df <- data.frame(type = c("advising","not advising"), number = c(10268,2443)) %>% mutate(ymax = cumsum(number), ymin = cumsum(number) - number, share = number/ sum(number))

df %>% ggplot(aes(x = 3.5, y = ymax)) +
  geom_rect(aes(xmin = 3, xmax = 4, ymin = ymin, ymax = ymax, fill = type)) +
  geom_label(aes(label = paste0(round(share * 100),"%"))) +
  coord_polar(theta = "y") +
  xlim(1,4) +
  theme_void() +
  theme(legend.position = "top") +
  scale_fill_manual(values = c("orange","red")) +
  labs(caption = "")
```

## b. The number of student per semester

```{r semester, fig.width= 15, fig.height= 8}
# Semester period 
student_semester<- student_new %>% distinct(identifier, .keep_all = T) %>% group_by(term_code) %>% tally() %>%  arrange(parse_number(term_code))
advising_semester <- advising_new %>% distinct(identifier, .keep_all = T) %>% group_by(term_code) %>% tally() %>% arrange(parse_number(term_code))

semester_table <- left_join(student_semester, advising_semester, by = "term_code", suffix = c("_student","_advising")) %>% gather(key = "type", value = "number", -c(term_code))

semester_table %>% ggplot(aes(x = term_code, y = number)) +
  geom_col(aes(fill = type), position = "dodge") +
  geom_label(aes(label = number, color = type, group = type), size = 4, position = position_dodge(width = 1)) +
  facet_wrap(parse_number(term_code) ~ ., scales = "free", ncol = 3) +
  coord_flip() +
  theme_bw() +
  labs(x = "", y = "Count") +
  labs(caption = "")
```

- On yearly average, Winter Semester has more student than Spring & Fall semesters.
- The students are less likely to visit the adivising center for advising over time.

## c. First attended Semster distribution

```{r first_attended, fig.width= 8}
# First attended semster

first_attended <- student_new %>% group_by(first_attended_term_code) %>% tally() %>%  arrange(parse_number(first_attended_term_code))

first_attended %>% ggplot(aes(x = first_attended_term_code, y = n)) +
  geom_col(aes(fill = reorder(first_attended_term_code, -parse_number(first_attended_term_code)))) +
  geom_label(aes(label = paste0(n," -",round(n/sum(n),2) * 100, "%"))) +
  labs(title = "Attended Semester", x= "Semster", y = "Count") + 
  theme_light() +
  theme(legend.position = "none") +
  coord_flip()
```

- We can see that this is the record of students who attended BYU-I on 2013 to 2014 season. 

# 4. Feature Engineering

Since this analysis aims to know the effect of advising on the student's retention or withdrawal rate, I would like to combine two dataset with joining key to see the data at once. We will use $left\_join$ syntax to contains unmatched observations which stands for __the student who has not been advised by advising center__.

## - Student_new & Advising_new Joining key

```{r joining}
joining_key <- student_new[,which(colnames(student_new) %in% colnames(advising_new))] %>% colnames()

advising_student <- 
  left_join(advising_new, student_new, by = joining_key)

# datatable(advising_student %>% head(100), options = list(scrollX = T))
```

## - Enrollment Status (Target)

- Since the objective of this analysis is to predict if the student graduated or not, "Withdrawn" & "No show" are relabeled as "Not Graduated". 

__Before__
```{r enrollment}
advising_student %>% group_by(enrollment_status) %>% tally() 
```

__After__
```{r enrollment_2}
advising_student <- advising_student %>% mutate(enrollment_status = ifelse(str_detect(enrollment_status,"Graduated"), "Graduated", "Not Graduated"))

advising_student$enrollment_status %>% unique()
```

## - Get the freshman & sophomore record of each student

In terms of Data Science, Prediction is the process of estimating the consequences (Target) out of past data. The data might be biased if data has not been collected objectively or there could be confounding variables which can change the result dramatically. For predicting graduation of each student, The cohort analysis is required in terms of time series. I need the records of student within the first half to predict their last consequences. There are steps to configure the range of data below.   


1. Identify the classification of student by measuring the credits taken at BYU-I only

```{r classification}
advising_student <- advising_student %>% 
  mutate(sum_of_cumulative_campus_credit = as.numeric(sum_of_cumulative_earned_credits) - as.numeric(transfer_earned_credits))
```

Even if there is a $sum\_of\_cumulative_earned\_credits$ in the dataset, it conatains the transferred credits from other institution.
It needs to be modified by calculating the credits taken at BYU-I only.

2. divide the data set with two different level A(Assoicate)/B(Bachelor)

```{r classification_2}
associate_st_ad_f <- advising_student %>% 
  filter(level == "A" & cumulative_earned_credits <= 30)

bachelor_st_ad_fs <- advising_student %>% 
  filter(level == "B" & cumulative_earned_credits <= 60)
```

Depending on the __level__ *(Bachleor vs Associate)*, I assume that 30 credits and 60 credits are half of credits for graudation of each degree. In student manual, it is stated that 60 credits are required for getting associate, 120 credits are required for getting bachelor degree. 

3. bind two datasets to one dataset

```{r classification_3}
advising_student_fs <- rbind(associate_st_ad_f, bachelor_st_ad_fs)
```

4. Identify the proportion of target variable

```{r classification_4}
table(advising_student_fs$enrollment_status) %>% data.frame() %>% mutate(Pct = round(Freq / sum(Freq) * 100, 2))
```


5. Categorize the Freshman ~ Senior with campus earned credits

```{r classification_5}
advising_student_fs <- 
advising_student_fs %>% 
  mutate(classification = case_when(
    cumulative_earned_credits <= 30 ~ "Fresh",
    cumulative_earned_credits <= 60 ~ "Sopho",
    cumulative_earned_credits <= 90 ~ "Junior",
    TRUE ~ "Senior"
  ))
```

## Sum of missing credits which were taken at BYU-I

```{r missing_credit}
advising_student_fs <-
advising_student_fs %>% 
  mutate(sum_of_cumulative_missing_credit = as.numeric(sum_of_cumulative_attempted_credits) + as.numeric(transfer_earned_credits) - as.numeric(sum_of_cumulative_earned_credits))
```

# Mean of missing Credits by enrollment status

```{r missing_credits}
t1 <- advising_student_fs %>% 
  group_by(enrollment_status) %>% 
  summarise(mean = mean(sum_of_cumulative_missing_credit, na.rm = T)) 

advising_student_fs %>% 
  ggplot(aes(x = enrollment_status, y = sum_of_cumulative_missing_credit)) +
  geom_jitter(alpha = 0.3, color = "gray") +
  geom_violin(aes(fill = enrollment_status)) +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip() +
  annotation_custom(tableGrob(t1, rows = NULL), xmin= 0.5, xmax=  1 , ymin= 50 , ymax= 60)
  
```

## - Mission_date 
### (Lubridate::%--% = as.interval())

```{r mission_date}
fix_year <- function(x, year=1919){ #since this year is 2019, if year digits are over 19, it is regared as 1900 century.
  m <- year(x) %% 100
  year(x) <- ifelse(m > year %% 100, 1900+m, 2000+m)
  x
}

advising_student_fs <-
advising_student_fs %>% 
  mutate(mission_enter_date = mdy(mission_enter_date) %>% fix_year(), mission_release_date = mdy(mission_release_date) %>% fix_year()) %>% 
  mutate(served_mission = mission_enter_date %--% mission_release_date %>% as.duration()/ ddays(1)) %>% 
  mutate(served_mission = case_when(
    is.na(served_mission) ~ "N",
    served_mission < 100 ~ "H",
    TRUE ~ "F"
  ))
```

- There are three categories for $served\_mission$ 
  * N represents for those who have not served mission yet
  * H represents for those who served mission but not completed in full period
    (if missionary came back within 100 days, I identify them as H)
  * F represents for those who served mission in full time

## - Act Score

I will only use $act\_composite$ for analysis so that we can reduce the dimesions of variable.

## - Birth Date switching to age and age group

```{r age}
advising_student_fs <- 
advising_student_fs %>% 
  mutate(age = round(mdy(birth_date) %>% fix_year() %--% ymd(paste0(20,parse_number(term_code)), truncated = 2L) %>% as.duration() / dyears(1))) %>% 
  mutate(age_group = case_when(
    age > 10 & age < 20 ~ "Teen",
    age < 40 ~  "Young Adult",
    age < 60 ~  "Mid Age",
    TRUE ~ "Aged"
  ))

t1 <- table(advising_student_fs$age_group) %>% data.frame() %>% tableGrob(rows=NULL)

advising_student_fs %>% ggplot(aes(x = age_group, y = age)) +
  geom_jitter(alpha = 0.5, color = "gray") +
  geom_violin(aes(fill = age_group), show.legend = FALSE) +
  annotation_custom(t1, ymin = 50, ymax = 80, xmin = 3.5, xmax =4.5) +
  scale_y_continuous(breaks = seq(10,80,10)) +
  theme_minimal()
```

## - Subprogram (Dayschool VS Online)

```{r subprogram}
advising_student_fs <-
advising_student_fs %>% 
  mutate(subprogram = substring(subprogram,1,1))
```

- Students can be classified into in-school student (D) or online student. (O)

## - Check if student transferred from other institution

```{r transfer}
advising_student_fs <-
advising_student_fs %>% 
  mutate(transfer = ifelse(transfer_earned_credits > 0, 1, 0) %>% as.factor())

advising_student_fs %>% group_by(transfer) %>% count()
```


## - Top 5 Major by level(degree) 

```{r major}
major_fs <- advising_student_fs %>% group_by(level, major) %>% tally(sort = T)

major_fs %>% group_by(level) %>% top_n(5) %>% ungroup() %>% 
  ggplot(aes(x = parse_number(major) %>% as.factor() %>% reorder(n), y = n)) +
  geom_col(aes(fill = level)) +
  geom_label(aes(label = str_extract(major, "[:alpha:]+"))) +
  facet_wrap(level ~., scales = "free", ncol = 1) +
  coord_flip() +
  theme_bw() +
  labs(x = "Major", y = "Count", title = "Top 5 Major by degree") +
  scale_fill_discrete(label = c("Assocaite", "Bachelor"))
```

## - Top 5 Advising Reason by level(degree)

```{r advising_reason, fig.height= 6, fig.width= 12}
advising_reason_fs <-
  advising_student_fs %>% group_by(level, advising_reason, enrollment_status) %>% count() %>% ungroup() %>% group_by(level, enrollment_status) %>% top_n(5, n)

advising_reason_fs %>% ggplot(aes(x = reorder(advising_reason, -n), y = n)) +
  geom_col(aes(fill = level), position = "dodge") +
  facet_wrap( level ~ enrollment_status, scales = "free") +
  theme_light() +
  coord_flip() +
  theme(legend.position = "top") +
  labs(x = "", y = "") +
  scale_fill_discrete(labels = c("Associate","Bachelor")) +
  labs(caption = "We can see that Graduated Students are more likely to visit advising center with the reason of Glad Planner & Class Planning compared to others")
```


## - Aggregate the data group by student id (Identifier)

After feature engineering, I aggregated the data by distinct Identifier to manage the individual data of each student.

```{r aggregation}
advising_student_final <- 
advising_student_fs %>% 
  group_by(identifier) %>% 
  arrange(cumulative_earned_credits) %>% 
  summarise(total_meeting_duration = sum(total_meeting_duration),
            number_of_semester = length(enrolled_current_semester[unique(term_code)] %>% str_detect("T")) ,
            classification_begin = head(classification_begin,1),
            served_mission = head(served_mission,1),
            subprogram = head(subprogram,1),
            gender = head(gender, 1),
            act_composite = mean(act_composite),
            high_school_gpa = mean(high_school_gpa),
            level = head(level, 1),
            cumulative_gpa = tail(cumulative_gpa, 1),
            transfer = head(transfer, 1),
            age = round(mean(age)),
            cumulative_earned_credits = head(cumulative_earned_credits, 1),
            cumulative_attempted_credits = head(cumulative_attempted_credits, 1),
            cumulative_missing_credit = round(max(cumulative_attempted_credits) - (max(cumulative_earned_credits) - head(transfer_earned_credits,1)),2),
            cumulative_campus_earned_credit = max(cumulative_earned_credits) - head(transfer_earned_credits,1),
            transfer_earned_credits = mean(transfer_earned_credits, na.rm = T),
            # target variable
            enrollment_status = head(enrollment_status, 1)) %>% 
  ungroup()

# datatable(advising_student_final,)
```


# 5. Modeling


## 1. Target Distribution

```{r taget}
proportion <- advising_student_final %>% group_by(enrollment_status) %>% count() %>% ungroup() %>%  mutate(pct = round(n/sum(n),2))
proportion
```

- `r proportion$pct[1] * 100`% of students in the data are composed of not graudated.
- `r proportion$pct[2] * 100`% of students are gradudated

## 2. Identify the missing variables and impute the part (median)

```{r missing_impute, warning = F}
naniar::gg_miss_var(advising_student_final, show_pct = T)
# naniar::gg_miss_fct(advising_student_final, enrollment_status)

# advising_student_final %>% ggplot() +
#   geom_histogram(aes(x = act_composite, fill = enrollment_status)) +
#   facet_wrap(~ enrollment_status) +
#   theme_minimal()
# 
# advising_student_final %>% ggplot() +
#   geom_histogram(aes(x = high_school_gpa, fill = enrollment_status)) +
#   facet_wrap(~ enrollment_status) +
#   theme_minimal()

advising_student_final %>% group_by(enrollment_status) %>% summarise(act_median = median(act_composite, na.rm = T), hs_gpa_median = median(high_school_gpa, na.rm = T))

advising_student_final[advising_student_final$enrollment_status == "Graduated" & is.na(advising_student_final$act_composite),"act_composite"] <- 22
advising_student_final[advising_student_final$enrollment_status == "Not Graduated" & is.na(advising_student_final$act_composite),"act_composite"] <- 22
advising_student_final[advising_student_final$enrollment_status == "Graduated" & is.na(advising_student_final$high_school_gpa),"high_school_gpa"] <- 3.46
advising_student_final[advising_student_final$enrollment_status == "Not Graduated" & is.na(advising_student_final$high_school_gpa),"high_school_gpa"] <- 3.24


advising_student_final <- advising_student_final %>% mutate_if(is.character, as.factor) %>% mutate_if(is.logical, as.factor) %>% select(-c(identifier))

sapply(advising_student_final, class)

naniar::gg_miss_var(advising_student_final, show_pct = T)
```

```{r modeling_1}
set.seed(1000)
InTraining <- caret::createDataPartition(advising_student_final$enrollment_status, p = .7, list = F)
Train <- advising_student_final %>% slice(InTraining)
Test <- advising_student_final %>% slice(-InTraining)

ModelFit_1 <- train(enrollment_status ~ ., data = Train, method = "rf", importance = T)
ModelFit_2 <- train(enrollment_status ~ ., data = Train, method = "rpart")
```

```{r modeling_2, fig.height= 7}
varimp1 <- ggplot(varImp(ModelFit_1)) + geom_label(aes(label = paste0(round(Importance),"%"))) + theme_bw() + labs(title = "Random Forest") + theme(plot.title = element_text(hjust = 0.5))
varimp2 <- ggplot(varImp(ModelFit_1)) + geom_label(aes(label = paste0(round(Importance),"%"))) + theme_bw() + labs(title = "RPart") + theme(plot.title = element_text(hjust = 0.5))
grid.arrange(varimp1, varimp2, ncol = 1)
```

```{r modeling_3}
caret::confusionMatrix(Test$enrollment_status ,predict(ModelFit_1, Test))
caret::confusionMatrix(Test$enrollment_status ,predict(ModelFit_2, Test))
```


```{r modeling_4}
fancyRpartPlot(ModelFit_2$finalModel)
```

```{r modeling_5}
colSums(is.na(advising_student_final))

advising_student_final <- advising_student_final %>% mutate(pred = predict(ModelFit_1, advising_student_final))

confusionMatrix(advising_student_final$enrollment_status, advising_student_final$pred)
```